---
title: "Technically Correct Data Preparation"
author: "Saloomeh (Sally) Arshi"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pollutant Data Cleaning

This segment of the code is focused on: identifying the rows of missing data introduced as a result of uploading CSV files and removing them, replacing the -999 and 9999 (as provided by the Ontario Ministry of Environment data definition) with NAs. 

```{r, echo=FALSE}
rm(list = ls())

# Loading libraries

library(ggplot2)
library(dplyr)
library(tidyr)
library(imputeTS)
library(forecast)
```

```{r}
# Read Files
CO <- read.csv('/Users//SallyA/Documents/Capstone CKME/Data Files/CO.csv')
NO <- read.csv('/Users//SallyA/Documents/Capstone CKME/Data Files/NO.csv')
NO2 <- read.csv('/Users//SallyA/Documents/Capstone CKME/Data Files/NO2.csv')
NOX <- read.csv('/Users//SallyA/Documents/Capstone CKME/Data Files/NOX.csv')
O3 <- read.csv('/Users//SallyA/Documents/Capstone CKME/Data Files/O3.csv')
PM25 <- read.csv('/Users//SallyA/Documents/Capstone CKME/Data Files/PM25.csv')

# It appears that there are 9999.00 and -999.00 and NAs in the data for missing values

# Omitting all NAs that are in rows at the end of the data frames
CO <- na.omit(CO)
NO <- na.omit(NO)
NO2 <- na.omit(NO2)
NOX <- na.omit(NOX)
O3 <- na.omit(O3)
PM25 <- na.omit(PM25)

# Replace all 9999.00s and -999.00s with NA
CO[CO==9999] <- NA
CO[CO==-999] <- NA

NO[NO==9999] <- NA
NO[NO==-999] <- NA

NO2[NO2==9999] <- NA
NO2[NO2==-999] <- NA

NOX[NOX==9999] <- NA
NOX[NOX==-999] <- NA

O3[O3==9999] <- NA
O3[O3==-999] <- NA

PM25[PM25==9999] <- NA
PM25[PM25==-999] <- NA

# Remove the first column from all dataframes as it is no necessary
CO$X = NULL
NO$X = NULL
NO2$X = NULL
NOX$X = NULL
O3$X = NULL
PM25$X = NULL
```

## Check Summary Statistics of Data
```{r}
summary(CO)
```
The rest of summary stats are supressed to reduce the size of the knitr file.
```{r eval=FALSE}
summary(NO)
summary(NO2)
summary(NOX)
summary(O3)
summary(PM25)
```

# Transpose the hourly data
First need to transpose all hourly data into rows in order to be able to build a time series class of data.
```{r}
# Transposing the hourly data
New_CO <- CO %>%
  pivot_longer(-Date, names_to = "Hour", values_to = "CO")
New_NO <- NO %>%
  pivot_longer(-Date, names_to = "Hour", values_to = "NO")
New_NO2 <- NO2 %>%
  pivot_longer(-Date, names_to = "Hour", values_to = "NO2")
New_NOX <- NOX %>%
  pivot_longer(-Date, names_to = "Hour", values_to = "NOX")
New_O3 <- O3 %>%
  pivot_longer(-Date, names_to = "Hour", values_to = "O3")
New_PM25 <- PM25 %>%
  pivot_longer(-Date, names_to = "Hour", values_to = "PM25")

# Convering Dates back to Date format
New_CO$Date <- as.Date(New_CO$Date)
New_NO$Date <- as.Date(New_NO$Date)
New_NO2$Date <- as.Date(New_NO2$Date)
New_NOX$Date <- as.Date(New_NOX$Date)
New_O3$Date <- as.Date(New_O3$Date)
New_PM25$Date <- as.Date(New_PM25$Date)
```

## Plotting the data

```{r pollutants, echo=FALSE}
#Plotting the Pollutants
g_CO <- ggplot(New_CO, aes(x=Date, y=CO))
p_CO <- g_CO + geom_line() + scale_x_date(date_labels = "%Y", date_breaks = "1 year") + theme_minimal()
print(p_CO)

g_NO <- ggplot(New_NO, aes(x=Date, y=NO))
p_NO <- g_NO + geom_line() + scale_x_date(date_labels = "%Y", date_breaks = "1 year") + theme_minimal()
print(p_NO)

g_NO2 <- ggplot(New_NO2, aes(x=Date, y=NO2))
p_NO2 <- g_NO2 + geom_line() + scale_x_date(date_labels = "%Y", date_breaks = "1 year") + theme_minimal()
print(p_NO2)

g_NOX <- ggplot(New_NOX, aes(x=Date, y=NOX))
p_NOX <- g_NOX + geom_line() + scale_x_date(date_labels = "%Y", date_breaks = "1 year") + theme_minimal()
print(p_NOX)

g_O3 <- ggplot(New_O3, aes(x=Date, y=O3))
p_O3 <- g_O3 + geom_line() + scale_x_date(date_labels = "%Y", date_breaks = "1 year") + theme_minimal()
print(p_O3)

g_PM25 <- ggplot(New_PM25, aes(x=Date, y=PM25))
p_PM25 <- g_PM25 + geom_line() + scale_x_date(date_labels = "%Y", date_breaks = "1 year") + theme_minimal()
print(p_PM25)
```

It is clear from these graphs that there are missing data especially in CO dataset where a chunk of data is missing. The missing data needs to be identified and imputed.
# NA Statistics 
To understand what imputation technique would work best. In order to do so, the statNA from ImputeTS package is leveraged to provide a summary statistics on the missing data.
```{r}
#Statistics on NAs
t_CO <- New_CO$CO 
t_NO <- New_NO$NO
t_NO2 <- New_NO2$NO2
t_NOX <- New_NOX$NOX
t_O3 <- New_O3$O3
t_PM25 <- New_PM25$PM25

statsNA(t_CO)
statsNA(t_NO)
statsNA(t_NO2)
statsNA(t_NOX)
statsNA(t_O3)
statsNA(t_PM25)
```
```{r gap size, echo=FALSE}
#Gap size plot
par(mfrow=c(1,1))
plotNA.gapsize(t_CO)
plotNA.gapsize(t_NO)
plotNA.gapsize(t_NO2)
plotNA.gapsize(t_NOX)
plotNA.gapsize(t_O3)
plotNA.gapsize(t_PM25)
```

As indicated there are many missing data which needs to be addressed by imputation. Imputation in time series is handled by converting the data to time series and leveraging the ImputeTS package capabilities. For this purpose interpolation is leveraged and of the three different types of interpolations (spline, linear and stine) the linear option is selected. This choice was based on chacking three different options and the summary stats of the dataset after imputation to see which one has the lowest impact on mean and median of the dataset. It was concluded that the linear option would best the best.

```{r}
#NA Interpolation Impute Linear, Spline, vs Stinterp
intp_CO <- na_interpolation(t_CO, option = "linear")
intp_NO <- na_interpolation(t_NO, option = "linear")
intp_NO2 <- na_interpolation(t_NO2, option = "linear")
intp_NOX <- na_interpolation(t_NOX, option = "linear")
intp_O3 <- na_interpolation(t_O3, option = "linear")
intp_PM25 <- na_interpolation(t_PM25, option = "linear")
```

# Post Imputaiton Plot of Technically Correct Data

Below graphs are the data post imputation. The data is now ready for exploratory data analysis.

```{r Technically Correct Data, echo=FALSE}
#Plotting imputed datasets - hourly time series to check imputation
TS_IMCO <- ts(intp_CO, start = c(2006, 1), end = c(2010, 12), frequency = 24*365)
plot(TS_IMCO)
```
The rest of graphs suppressed due to the size of knitr file.
```{r eval=FALSE}
TS_IMNO <- ts(intp_NO, start = c(2006, 1), end = c(2010, 12), frequency = 24*365)
plot(TS_IMNO)
TS_IMNO2 <- ts(intp_NO2, start = c(2006, 1), end = c(2010, 12), frequency = 24*365)
plot(TS_IMNO2)
TS_IMNOX <- ts(intp_NOX, start = c(2006, 1), end = c(2010, 12), frequency = 24*365)
plot(TS_IMNOX)
TS_IMO3 <- ts(intp_O3, start = c(2006, 1), end = c(2010, 12), frequency = 24*365)
plot(TS_IMO3)
TS_IMPM25 <- ts(intp_PM25, start = c(2006, 1), end = c(2010, 12), frequency = 24*365)
plot(TS_IMPM25)
```
